Semantic Linearity in a Lazy Optimising Compiler

### 2025-02-23 icfp.tex
Up to 3.6.2

%Case expressions \emph{drive evaluation} --
% thus they are key to realize a type system that understands linearity in the presence of lazy evaluation.
%
%However, the evaluation of case expressions is considerably nuanced,
%
%In lazy let bindings, computations  can be a case
%expression can effectively consume resources rather than just 
%
A case expressions drive evaluation by evaluating their scrutinee to Weak Head
Normal Form (WHNF)~\cite{10.5555/1096899}. Then, the case
alternative whose pattern matches the result of evaluating the
scrutinee is taken\footnote{In our calculus, the alternatives are assumed to be
  exhaustive, i.e. there always exists at least one pattern which
  matches the scrutinee in its WHNF, so we're guaranteed to have an
  expression to progress evaluation.}.
An expression in WHNF can either be %\begin{itemize}
% \item
a $\lambda$-abstraction $(\lambda x.~e)$ 
or a datatype constructor application $(K~\ov{e})$.
%\end{itemize}
In both cases, the sub-expressions $e$ or $\ov{e}$ occurring in the lambda body
or as constructor arguments need not be evaluated for the lambda or constructor
application to be in WHNF. % (if otherwise all sub-expressions
% were fully evaluated the whole expression would also be in \emph{normal form}).
% (that is why it is called \emph{weak} head normal form).
%
Accordingly, these sub-expressions might still depend on linear resources to be
well-typed (such resources will only be consumed when the sub-expressions themselves are evaluated).

@@

We highlight that, when evaluating a case expression, computation only
happens when a scrutinee (not in WHNF) is evaluated to
WHNF. Otherwise, evaluation continues in the alternative by
substituting in the appropriate scrutinee expressions, but without
having performed any additional computation.
% (the scrutinee was already in weak head normal form).
%In short, no computation happens if the scrutinee is already in WHNF.
%
In terms of linearity, resources are consumed only if evaluation happens.
Therefore, resources used to type a scrutinee not in
WHNF will be consumed when the case is evaluated, making said resources unavailable in the case
alternatives. Conversely, when the scrutinee is already in WHNF, linear
resources required to type the scrutinee are still available in the alternatives.
The linear resources used by an expression in WHNF are exactly those which
occur to the right of $\gtrdot$ in the WHNF judgement shown above
(corresponding to the resources required to typecheck the lambda body or the
constructor arguments).

@@

The dichotomy between evaluation (i.e.,~resource usage) of a case expression
whose scrutinee is in weak head normal form, or otherwise, leads to one of our
key insights: we must \emph{branch} on weak head normal formed-ness to
accurately type case expressions.
%
When the scrutinee is already in WHNF, the resources are
unused upon evaluation and thus available in the alternatives.
%
When it is not, resources will be consumed and cannot be used in the
alternative.
%
To illustrate, consider the following case expressions:
\[
\begin{array}{ccc}
(1)~\lambda x.~\ccase{K~x}{\_ \to x} &  & (2)~\lambda x.~\ccase{free~x}{\_ \to x}
\end{array}
\]
The first function uses $x$ linearly, while the second does not.
%
Alternatives may also use the case binder or pattern variables, referring to, respectively,
the whole scrutinee (and all resources used to type the scrutinee) or
constructor arguments (and the resources to type each argument).
% Using the case binder entails using all the resources required by the scrutinee, and using a pattern
% variable implies using the resources of the corresponding constructor argument.

There are three \emph{competing} ways
to use the resources from a scrutinee in WHNF in a case alternative: directly, via
the case binder, or by using \emph{all} the pattern-bound variables.
%
Recall how $\D$-variables can encode mutual exclusivity between alternative
ways of consuming resources -- it follows that case binders and pattern-bound
variables are another instance of $\D$-bound variables. Intuitively, resources
in a scrutinee that is already in WHNF are only properly consumed when all
(linear) fields of the pattern are used, satisfying the definition of
consuming resources given in Linear Haskell.
%
% We type a case expression whose scrutinee is in weak head normal form with
This suggests the following first attempt at a rule for cases of
scrutinees in WHNF:
\[
\TypeCaseWHNFIntermediate
\]
First, we assert this rule is only applicable to expressions in weak head
normal form. Second, we appeal to the typing judgement for expressions in WHNF
to determine the split of resources amongst the scrutinee
sub-expressions. Finally, we type all case alternatives with the same context, using the
$\vdash_{alt}$ judgement, introducing
the case binder $z$ in the environment as a $\D$-bound
variable whose usage environment is the linear resources used to type the
scrutinee. We also 
make all the resources $\ov{\D_i}$ used to type the scrutinee
available in the linear typing environment.
%
The $\vdash_{alt}$ judgement is 
annotated with the disjoint set of linear resources $\ov{\D_i}$ used
to typecheck the scrutinee sub-expressions,
and the name of the case binder $z$.

%
Despite the key intuitions of typing a case expressions whose scrutinee is in
WHNF being conveyed by this rule, our treatment of such a case expression is
slightly more involved, as will be presented after discussing cases of
scrutinees not in WHNF.

The alternative judgement $\G;\D \vdash_{alt} \rho \to e :^z_\D \s \Rightarrow
\vp$ is used to type case alternatives, encompassing three
``modes'', distinguished by the kind of arrow that is used:
for alternatives of case expressions whose scrutinee is in WHNF ($\Rightarrow_\mathsf{WHNF}$),
for case expressions in which the scrutinee is not in WHNF ($\Rightarrow_\mathsf{NWHNF}$),
and for alternatives agnostic to the WHNF-ness of the scrutinee
($\Rightarrow$), with $\Rightarrow$ also generalizing the other two.
%
% Following the $Case_\textrm{WHNF}$ rule in which we use the $\Mapsto$ alternative
% judgement, t
The rule for typing an alternative whose pattern is a
constructor with $n > 0$ linear components (in mode $\Rightarrow_{\mathsf{NWHNF}}$) is:
\[
\TypeAltNWHNF
\]
The rule states that, for such a pattern matching a scrutinee already in
WHNF, we introduce the linear components of the pattern as $\D$-bound variables
whose usage environment matches the linear resources required to type the
corresponding constructor argument in the scrutinee, which is annotated in
the judgement ($\ov{\D_i}$). Unrestricted fields of the constructor are
introduced as unrestricted variables. We note that the typing environment $\D$
always contains the resources $\ov{\D_i}$ in uses of the alternative
judgement.

The rule for alternatives that match on the wildcard pattern is:
\[
\TypeAltWild
\]
To type a wildcard alternative we simply type the expression with the main
judgement, ignoring all annotations on the judgement. We recall that the case
binder was already introduced in the environment with the appropriate usage
environment by the case expression.

Finally, consider an alternative matching on a case constructor without any
linear components. According to the definition of consuming a resource from
Linear Haskell, the linear resources of a scrutinee matching such a pattern are
fully consumed in the body of the corresponding alternative, since the
scrutinee must have been evaluated to a form that does not have any linear
components.
%
This definition agrees with the intuition we have developed by example in the
previous section, and with the typing rule we devised for alternatives matching
constructors without linear components.
Taking into account that case expressions introduce the linear resources of the
scrutinee in the typing environment of all alternatives, and in the usage
environment of the case binder, we must reactively update the typing
environments after matching on such a pattern.
%
The $Alt0$ rule essentially encodes this insight, and is applicable regardless
of the WHNF-ness of the scrutinee ($\Rightarrow$ mode), as long as
the constructor pattern has no linear fields:
%
\[
\TypeAltZero
\]
The rule deletes the annotated scrutinee environment $\D_s$ from
%two select environments:
the linear typing context (written $\D[\cdot/\D_s]$, a substitution of
the scrutinee typing environment by the empty linear environment $\cdot$)
and from the usage environment of the case binder $z$, written $\G[\cdot/\D_s]_z$
to denote replacing the usage environment of the variable $z$ in $\G$, which
is necessarily $\D_s$ (since we always annotate the judgement with the
environment of the scrutinee), by the empty environment.

@@

% \end{itemize}
%
The rule faithfully encodes the notion that an expression matching such a
pattern is unrestricted when evaluated to WHNF, implying that all linear
resources have been consumed to produce it, and the result may 
be freely discarded or duplicated.
%
Reactively deleting the resources that typed the scrutinee from the linear
environment ensures that when we match on an unrestricted pattern there are no
remaining linear resources to consume. Otherwise, e.g., $\ccase{K_1~x}{\{K_2 \to
K_2,K_1~y \to K_1~y\}}$ would not be well-typed since the resource $x$ would
be left \emph{un}-consumed in the first alternative, even though we know by
matching on $K_1$ that it has already been consumed.
%
Furthermore, since the case binder in such an alternative refers to the
unrestricted expression, the case binder too may be used unrestrictedly, which
we allow by emptying its usage environment.

@@

We thank Arnaud Spiwack for insightful discussions and Well-Typed for funding
this work.\todo{Do these have to be anonymized?}

>>>>

%Case expressions \emph{drive evaluation} --
% thus they are key to realize a type system that understands linearity in the presence of lazy evaluation.
%
%However, the evaluation of case expressions is considerably nuanced,
%
%In lazy let bindings, computations  can be a case
%expression can effectively consume resources rather than just 
%
A case expression drives evaluation by evaluating its scrutinee to Weak Head
Normal Form (WHNF)~\cite{10.5555/1096899}. Then, the case
alternative whose pattern matches the result of evaluating the
scrutinee is taken\footnote{In our calculus, the alternatives are assumed to be
  exhaustive, i.e. there always exists at least one pattern which
  matches the scrutinee in its WHNF, so we're guaranteed to have an
  expression to progress evaluation.}.
An expression in WHNF can either be %\begin{itemize}
% \item
a $\lambda$-abstraction $(\lambda x.~e)$ 
or a datatype constructor application $(K~\ov{e})$.
%\end{itemize}
In both cases, the sub-expressions $e$ or $\ov{e}$ occurring in the lambda body
or as constructor arguments need not be evaluated for the lambda or constructor
application to be in WHNF. % (if otherwise all sub-expressions
% were fully evaluated the whole expression would also be in \emph{normal form}).
% (that is why it is called \emph{weak} head normal form).
%
Accordingly, these sub-expressions might still depend on linear resources to be
well-typed (such resources will only be consumed when the sub-expressions themselves are evaluated).

@@

We highlight that, when evaluating a case expression, computation only
happens when a scrutinee (not in WHNF) is evaluated to
WHNF. Otherwise, evaluation continues in the alternative by
substituting in the appropriate scrutinee expressions, but without
having performed any additional computation.
% (the scrutinee was already in weak head normal form).
%In short, no computation happens if the scrutinee is already in WHNF.
%
In terms of linearity, resources are consumed only if evaluation happens.
Therefore, resources used to type a scrutinee not in
WHNF will be consumed when the case is evaluated and no longer be directly available in the case
alternatives. Conversely, when the scrutinee is already in WHNF, linear
resources required to type the scrutinee are still directly available in the
alternatives.
%
%%%%???
%%%%The linear resources used to type an expression in WHNF are exactly those which
%%%%occur to the right of $\gtrdot$ in the WHNF judgement shown above
%%%%(corresponding to the resources required to typecheck the lambda body or the
%%%%constructor arguments).

@@

The dichotomy between evaluation (and resource usage) of a case expression
whose scrutinee is in weak head normal form and one whose scrutinee is not
leads to one of the key insights of $\lambda^\pi_\Delta$: we must \emph{branch}
on weak head normal formed-ness to accurately type case expressions.
%
When the scrutinee is already in WHNF, the resources are
unused upon evaluation and thus available in the alternatives.
%
When it is not, resources will be consumed and cannot be used in the
alternative.
%
To illustrate, consider the following case expressions:
\[
\begin{array}{ccc}
    (1)~\lambda x.~\ccase{K~x}{z~\{\_ \to x\}} &  & (2)~\lambda x.~\ccase{free~x}{z~\{\_ \to x\}}
\end{array}
\]
The first function uses $x$ linearly, but the second does not.
%
Alternatives may also use the case binder or pattern variables, referring to, respectively,
the whole scrutinee (and all resources used to type the scrutinee) or
constructor arguments (and the resources to type each argument).
% Using the case binder entails using all the resources required by the scrutinee, and using a pattern
% variable implies using the resources of the corresponding constructor argument.

There are three competing ways
to use the resources from a scrutinee in WHNF in a case alternative: directly, via
the case binder, or by using pattern-bound variables.
%
Recall how $\D$-variables can encode mutual exclusivity between alternative
ways of consuming resources -- it follows that case binders and pattern-bound
variables are yet another instance of $\D$-bound variables.
%
%Resources in a scrutinee that is already in WHNF are only consumed when all
%(linear) fields of the pattern are used, satisfying the definition of
%consuming resources given in Linear Haskell.
%
% We type a case expression whose scrutinee is in weak head normal form with
This suggests the following first rule for cases of scrutinees already in WHNF:
\[
\TypeCaseWHNFIntermediate
\]

First, we assert this rule is only applicable to expressions in weak head
normal form. Second, we appeal to the typing judgement for expressions in WHNF
to determine the split of resources amongst the scrutinee
sub-expressions. Finally, we type all case alternatives with the same context, using the
$\vdash_{alt}$ judgement, introducing
the case binder $z$ in the unrestricted environment as a $\D$-bound
variable whose usage environment is the linear resources $\ov{\D_i}$ used to type the
scrutinee. Those same resources $\ov{\D_i}$ are again made available in the
linear typing environment of the alternative, similarly to how the resources
used to type a $Let$ binder are still available in the continuation.
%
%The $\vdash_{alt}$ judgement is 
%annotated with the disjoint set of linear resources $\ov{\D_i}$ used
%to type the scrutinee sub-expressions,
%and the name of the case binder $z$.

%
Despite the main idea for typing ``WHNF case expressions'' being conveyed by
this rule, in our system it is slightly more involved, as will be seen after
discussing cases of scrutinees not in WHNF.

The alternative judgement $\G;\D \vdash_{alt} \rho \to e :^z_\D \s \Rightarrow
\vp$ is used to type case alternatives, encompassing three
``modes'' distinguished by the kind of arrow that is used:
for alternatives of case expressions whose scrutinee is in WHNF
($\Rightarrow_\mathsf{WHNF}$), not in WHNF ($\Rightarrow_\mathsf{NWHNF}$),
and for alternatives agnostic to the WHNF-ness of the scrutinee
($\Rightarrow$), with $\Rightarrow$ also generalizing the other two.
%
% Following the $Case_\textrm{WHNF}$ rule in which we use the $\Mapsto$ alternative
% judgement, t

First, an alternative whose pattern is a constructor with $n > 0$ linear components
under the $\Rightarrow_{\mathsf{NWHNF}}$ mode is typed as:
\[
\TypeAltNWHNF
\]
The rule states that, for such a match on a scrutinee already in
WHNF, we introduce the linear components of the pattern as $\D$-bound variables
with usage environment matching the linear resources required to type the
corresponding constructor argument in the scrutinee. The resources required to
type each constructor sub-expression come annotated in the judgement (as
${:}_{\ov{\D_i}}$). Unrestricted fields of the constructor are introduced as
unrestricted variables. We note that the typing environment $\D$ always
contains the resources $\ov{\D_i}$ in uses of the alternative judgement.

Second, the rule for alternatives that match on the wildcard pattern, be the
original scrutinee in WHNF or not (note the use of $\Rightarrow$, which
is applicable both under $\Rightarrow_{WHNF}$ and $\Rightarrow_{NWHNF}$):
\[
\TypeAltWild
\]
To type a wildcard alternative we simply type the expression with the main
judgement, ignoring all annotations on the judgement. The case binder will
have already been introduced in the environment with the appropriate usage
environment by the relevant (WHNF or not) case expression rule.

Lastly, consider an alternative matching on a data constructor without any
linear components. The linear resources used to type a scrutinee matching such
a pattern are fully consumed during evaluation: the resulting unrestricted
constructor application can only be well-typed under an empty linear
environment.
%
Consequently, the resources to type the scrutinee which are carried over to the
case-alternative environments by the $Case_{WHNF}$ and $Case_{NWHNF}$ rules
(see the latter below), must be reactively destroyed from branches where the
pattern is unrestricted. In failing to do this, we'd allow the fully-consumed
resources to be used again in the branch!
% Really, in the WHNF case it is not really necessary to remove the things
% since they must already be empty be definition.
%
%This definition agrees with the intuition we have developed by example in the
%previous section, and with the typing rule we devised for alternatives matching
%constructors without linear components.
%
%Taking into account that case expressions introduce the linear resources of the
%scrutinee in the typing environment of all alternatives, and in the usage
%environment of the case binder, we must reactively update the typing
%environments after matching on such a pattern.
%
The $Alt0$ rule essentially encodes this insight, and is applicable regardless
of the WHNF-ness of the scrutinee ($\Rightarrow$ mode), as long as
the constructor pattern has no linear fields:
%
\[
\TypeAltZero
\]
The rule deletes the annotated scrutinee environment $\D_s$ from
%two select environments:
the linear typing context (written $\D[\cdot/\D_s]$, a substitution of
the scrutinee typing environment by the empty linear environment $\cdot$)
and from the usage environment of the case binder $z$ (written $\G[\cdot/\D_s]_z$
to similarly delete $\D_s$ from the usage environment of the $\D$-var $z$ in $\G$).
Given that the case-binder $z$ is always annotated with $\D_s$ (essentially,
the linear environment typing the scrutinee), this latter substitution makes
the case-binder unrestricted in branches with unrestricted patterns.
%-- agreeing with our observation in Section \label{sec:semantic-linearity-examples}.

@@

% \end{itemize}
%
This rule observes that any linear resources required to type an expression
that reduces to an unrestricted constructor application are fully consumed
through evaluation and the resulting unrestricted expression can be freely
discarded or duplicated (e.g. via the case binder).
%
%Reactively deleting the resources that typed the scrutinee from the linear
%environment ensures that when we match on an unrestricted pattern there are no
%remaining linear resources to consume. Otherwise, e.g., $\ccase{K_1~x}{\{K_2 \to
%K_2,K_1~y \to K_1~y\}}$ would not be well-typed since the resource $x$ would
%be left \emph{un}-consumed in the first alternative, even though we know by
%matching on $K_1$ that it has already been consumed.
%
%Furthermore, since the case binder in such an alternative refers to the
%unrestricted expression, the case binder too may be used unrestrictedly, which
%we allow by emptying its usage environment.

@@

We thank Arnaud Spiwack for valuable and insightful discussions.\todo{Do these have to be anonymized?}
